High-Dimensional Regularized Discriminant Analysis

### 2016-10-23 reports/paper-hdrda.Rnw
Major rewrite of the intro. #28

As machine-learning algorithms are increasingly utilized to automate data-driven
decisions, researchers continue to develop a wealth of \emph{classifiers}
employed in a variety of disciplines, such as medical research, facial
recognition, remote sensing, and chemical analysis
\citep*{Ching:2012fu,Li:2012ev,Zhang:2010va}. Despite the advances in
classification to facilitate technological advancements, well-established
classifiers often degrade in accuracy when applied to modern, high-dimensional
data sets if the number of features $p$ exceeds the total sample size $N$, due
to the \emph{curse of dimensionality} \citep{Bellman:1961tn}. For instance,
linear discriminant analysis (\emph{LDA}) and quadratic discriminant analysis
(\emph{QDA}) are popular because of their simplicity, speed of computation, and
interpretability. However, both methods are incalculable when $p > N$ because
the class sample covariance matrices are singular \citep*{Bouveyron:2007gx}.

As a result, researchers have emphasized three standard approaches to modifying
the \emph{LDA} and \emph{QDA} classifiers for application to high-dimensional
data: 1) covariance-matrix regularization, 2) dimension reduction, and 3)
parameter restrictions. Covariance-matrix regularization methods are often based
on the well-known ridge-regression approach of \cite{Hoerl:1970cd}.
Additionally, \cite{Ching:2012fu}, \cite{Zhang:2010va}, \cite{Ye:2009gd},
\cite{Ji:2008wp}, \cite*{Guo:2007te}, \cite*{Srivastava:2007ud},
\cite{Ye:2006vx}, and \cite{Ye:2005uu} have proposed regularization methods to
improve the estimation of the class covariance matrices by adjusting the
eigenvalues of the sample covariance matrices to ensure positive
definiteness. Alternatively, dimension reduction typically transforms the
original data in the high-dimensional space to a subspace having a much smaller
dimension such that redundant and noisy information is removed
\citep*{Fan:2012iq,Ye:2009gd,Bouveyron:2007gx,Ye:2006vx,Ye:2005hd,Antoniadis:2003bk}.
Finally, \cite*{Mai:2012bf}, \cite{Fan:2012iq}, \cite*{Merchante:2012vk},
\cite*{Clemmensen:2011kr}, \cite*{Huang:2010ju}, and \cite*{Dudoit:2002ev} have
proposed restrictions on the population covariance matrices to reduce the number
of parameters that must be estimated. For instance, \cite{Dudoit:2002ev} have
assumed that the features within each class are independent so that the
population covariance matrices are diagonal, resulting in positive-definite
sample covariance matrices. For excellent overviews of the regularization
problem in discriminant analysis and other techniques, see \cite{Murphy:2012uq}
and \cite*{Mkhadri:1997gy}.

Here, we present the high-dimensional \emph{RDA} (\emph{HDRDA}) classifier,
which is intended for the $p > N$ case and builds on the classic \emph{RDA}
classifier proposed by \cite{Friedman:1989tm}. In this paper, we study the
classification performance of the \emph{HDRDA} classifier on six real
high-dimensional data sets along with a simulation design that generalizes the
experiments initially conducted by \cite{Guo:2007te}. We demonstrate that the
\emph{HDRDA} classifier often attains superior classification accuracy to
several recent classifiers designed for small-sample, high-dimensional data from
\cite*{Tong:2012hw}, \cite{Witten:2011kc}, \cite*{Pang:2009ik}, and
\cite{Guo:2007te}. We also include as a benchmark the random forest from
\cite{Breiman:2001fb} because \cite*{FernandezDelgado:2014ul} have concluded
that the random forest is often superior to other classifiers in benchmark
studies. We show that our proposed classifier is competitive and often
outperforms the random forest in terms of classification accuracy in the
small-sample, high-dimensional setting.

Our proposed \emph{HDRDA} classifier employs a biased covariance-matrix
estimator that partially pools the individual sample covariance matrices from
the \emph{QDA} classifier with the pooled sample covariance matrix from the
\emph{LDA} classifier. We then shrink the resulting covariance-matrix estimator
towards a scaled identity matrix to ensure positive definiteness. By
reparameterizing the \emph{RDA} classifier similar to that of
\cite*{Hastie:2008dt} and \cite{Halbe:2007ea}, we show that the pooling
parameter in the \emph{HDRDA} classifier determines the contribution of each
training observation to the estimation of each class covariance matrix, enabling
interpretability that has been previously lacking with the \emph{RDA} classifier
\citep{Bensmail:1996fw}.

Next, we achieve a substantial reduction in dimension that greatly reduces the
computational effort during model selection when $p \gg N$ by applying reasoning
similar to that of \cite{Ye:2006tq}. We establish that the matrix operations
corresponding to the null space of the pooled sample covariance matrix are
redundant and can be discarded from the \emph{HDRDA} decision rule without loss
of classificatory information. Moreover, because the \emph{HDRDA} decision rule
is invariant to adjustments to the zero eigenvalues, the decision rule in the
original feature space is equivalent to a decision rule in a lower dimension,
such that matrix inverses and determinants of relatively small matrices can be
rapidly computed. Also, we provide an efficient algorithm along with pseudocode
to estimate the \emph{HDRDA} classifier's tuning parameters in a grid search via
cross-validation. Finally, we show that several shrinkage methods that are
special cases of the \emph{HDRDA} classifier have no effect on the zero
eigenvalues of the covariance-matrix estimators. Such techniques include work
from \cite{Srivastava:2007ww}, \cite{Rao:1971ul}, and other methods studied by
\cite{Ramey:2013ji} and \citet*{Xu:2009fl}.

>>>>

In this paper, we consider the classification of small-sample, high-dimensional
data, where the number of features $p$ exceeds the training sample size $N$. In
this setting, well-established classifiers such as linear discriminant analysis
(\emph{LDA}) and quadratic discriminant analysis (\emph{QDA}) become
incalculable because the class and pooled covariance matrix estimators are
singular \citep*{Murphy:2012uq, Bouveyron:2007gx, Mkhadri:1997gy}. To improve
the estimation of the class covariance matrices estimated in the \emph{QDA}
classifier and to ensure the covariance matrix estimators are nonsingular,
\cite{Friedman:1989tm} proposed the regularized discriminant analysis
(\emph{RDA}) classifier by incorporating a weighted average of the pooled sample
covariance matrix and the class sample covariance matrix. To further improve the
estimation of the class covariance matrix and to stabilize its inverse,
\cite{Friedman:1989tm} also included a regularization component by shrinking the
covariance matrix estimator towards the identity matrix, which yields a
nonsingular estimator following the well-known ridge-regression approach of
\cite{Hoerl:1970cd}. Despite its popularity, the ``borrowing'' operation
employed in the \emph{RDA} classifier lacks interpretability
\citep{Bensmail:1996fw}, and the \emph{RDA} classifier is computationally
inefficient in that its matrix operations are cubic in the number of
features.\footnote{TODO: Add a couple of references saying RDA is slooooow.} As
a result, the \emph{RDA} classifier is impraticial for modern high-dimensional
data sets because its matrix operations are cubic in the number of features, and
the estimation of the \emph{RDA} classifier's two tuning parameters is
computationally burdensome.

Here, we present the high-dimensional \emph{RDA} (\emph{HDRDA}) classifier,
which is intended for the $p > N$ case and builds on the classic \emph{RDA}
classifier proposed by \cite{Friedman:1989tm}. We reparameterize the \emph{RDA}
classifier similar to that of \cite*{Hastie:2008dt} and \cite{Halbe:2007ea} and
employ a biased covariance-matrix estimator that partially pools the individual
sample covariance matrices from the \emph{QDA} classifier with the pooled sample
covariance matrix from the \emph{LDA} classifier. We then shrink the resulting
covariance-matrix estimator towards a scaled identity matrix to ensure positive
definiteness. We show that the pooling parameter in the \emph{HDRDA} classifier
determines the contribution of each training observation to the estimation of
each class covariance matrix, enabling interpretability that has been previously
lacking with the \emph{RDA} classifier \citep{Bensmail:1996fw}. Our
parameterization differs from \cite*{Hastie:2008dt} and \cite{Halbe:2007ea} in
that our formulation allows the flexibility of various covariance-matrix
estimators proposed in the literature, including a variety of ridge-like
estimators such as the one proposed by \cite{Srivastava:2007ww}.

Next, we establish that the matrix operations corresponding to the null space of
the pooled sample covariance matrix are redundant and can be discarded from the
\emph{HDRDA} decision rule without loss of classificatory information by
applying reasoning similar to that of \cite{Ye:2006tq}. As a result, we achieve
a substantial reduction in dimension such that the matrix operations used in the
\emph{HDRDA} classifier are computationally linear in the number of features.
Furthermore, we demonstrate that the \emph{HDRDA} decision rule is invariant to
adjustments to the approximately $p - N$ zero eigenvalues, so that the decision
rule in the original feature space is equivalent to a decision rule in a lower
dimension, such that matrix inverses and determinants of relatively small
matrices can be rapidly computed. Finally, we show that several shrinkage
methods that are special cases of the \emph{HDRDA} classifier have no effect on
the approximately $p - N$ zero eigenvalues of the covariance-matrix estimators
when $p > N$. Such techniques include work from \cite{Srivastava:2007ww},
\cite{Rao:1971ul}, and other methods studied by \cite{Ramey:2013ji} and
\citet*{Xu:2009fl}.

We provide an efficient algorithm along with pseudocode to estimate the
\emph{HDRDA} classifier's tuning parameters in a grid search via
cross-validation. Timing comparisons between our \emph{HDRDA} implementation in
the {\tt sparsediscrim} R package available on CRAN and the standard \emph{RDA}
formulation in the {\tt klaR} R package demonstrate that as the number of
features increases, the computational runtime of the \emph{HDRDA} classifier is
drastically smaller than \emph{RDA}. In fact, when $p = 1000$ and $N = 100$, the
\emph{HDRDA} classifier is \textbf{XXX} times faster on average than the
\emph{RDA} classifier.

Finally, we study the classification performance of the \emph{HDRDA} classifier
on six real high-dimensional data sets along with a simulation design that
generalizes the experiments initially conducted by \cite{Guo:2007te}. We
demonstrate that the \emph{HDRDA} classifier often attains superior
classification accuracy to several recent classifiers designed for small-sample,
high-dimensional data from \cite*{Tong:2012hw}, \cite{Witten:2011kc},
\cite*{Pang:2009ik}, and \cite{Guo:2007te}. We also include as a benchmark the
random forest from \cite{Breiman:2001fb} because \cite*{FernandezDelgado:2014ul}
have concluded that the random forest is often superior to other classifiers in
benchmark studies. We show that our proposed classifier is competitive and often
outperforms the random forest in terms of classification accuracy in the
small-sample, high-dimensional setting.
